Trait selection from within and between species variations while integrating genomics

### 2023-07-18 manuscript/main.tex
Updated manuscript. Ensure simulated variances are >0

\title{Trait' selection from within and between species variations taking in genomics}
\rhead{\scshape Test of trait selection}

@@

Even though the neutrality index can be applied for any trait, we so far assumed that several traits would to be independent of one another.
Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021}.
Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance-rejection), and the posterior distribution of $\CovarianceMatrix$ is sampled using a Gibbs sampler.
The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.

@@

We analyzed a mammalian dataset, with the body size and brain mass as quantitative traits.
The log-transformed values of body size and brain mass are gathered from \textcite{tsuboi_breakdown_2018}.
We removed individuals not marked as adult, and split the data into males and females due to sexual dimorphism.
We discarded species with no intra-specific variability.
The mammalian nucleotide diversity is obtained from the Zoonomia project, with nucleotide divergence obtained on a set of neutral markers in \textcite{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \textcite{wilder_contribution_2023}.

@@

The inference framework was first tested on independently simulated datasets.
With the aim of applying the inference method to empirical datasets, the simulation parameters were chosen so as to match an empirically relevant mammalian empirical regime (see Materials and Methods).
First, we assumed a constant population size ($\Ne$) and a constant mutation rate ($\MutationRate$) across the phylogeny (fig.~\ref{fig:results-simulations}A\&B)
Simulations performed under stabilizing selection where detected as such ($\proba [\EstNI < 1] > 0.95$), no false negative was observed (blue in fig.~\ref{fig:results-simulations}A\&B).
Simulations performed under diversifying selection were detected as such ($\proba [\EstNI > 1] > 0.975$), no false negative was observed (red in fig.~\ref{fig:results-simulations}A\&B).
Simulations performed under neutral evolution were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$) in 79\% of the cases (yellow in fig.~\ref{fig:results-simulations}A\&B), the remaining cases were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$) in 21\% of the simulations.
Second, for simulations under a fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}C\&D), simulations of diversifying selection was detected as such and stabilizing selection was detected as such in 100\% of the cases.
For simulations under neutral evolution, 46\% of the simulations were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), 54\% of the simulations were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$).

@@

\subsection*{Results on empirical data}\label{subsec:results-on-empirical-data}
For mammalian body size, we obtained males (\Male) trait and genetic variation for $\NbrTaxa=36$ species, and in females (\Female) variation for $\NbrTaxa=26$ species.
The posterior estimate of neutrality index are $\EstNI_{\text{\Male}} = 0.346$ ($[0.183,~0.677]$ for $95\%$ credible interval) and $\EstNI_{\text{\Female}} = 0.278$ ($[0.136,~0.639]$ for $95\%$ credible interval).
For mammalian brain mass, we obtained $\EstNI_{\text{\Male}} = 1.375$ ($[0.844,~2.212]$ for $95\%$ credible interval, $\NbrTaxa=36$) and $\EstNI_{\text{\Female}} = 1.762$ ($[1.021,~3.010]$ for $95\%$ credible interval, $\NbrTaxa=26$).
In both males and females, brain mass is under diversifying selection, with p-values of $0.116$ and $0.022$ respectively.

For primates body size, we obtained $\EstNI_{\text{\Male}} = 0.557$ ($[0.406,~0.767]$ for $95\%$ credible interval, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 0.393$ ($[0.277,~0.561]$ for $95\%$ credible interval, $\NbrTaxa=65$).
For primates brain mass, we obtained $\EstNI_{\text{\Male}} = 1.920$ ($[1.397,~2.616]$ for $95\%$ credible interval, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 1.974$ ($[1.393,~2.798]$ for $95\%$ credible interval, $\NbrTaxa=65$).
In both males and females, brain mass is under diversifying selection, with p-values of $0.00$ and $0.00$ respectively.
However, these tests are invalid because the assumption of neutral genetic markers used for normalization is violated (see table~\ref{table:assumptions}).

@@

Heritability ($h^2$) is defined as:
\begin{equation}
    h^2 = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}\label{eq:simu-heritability}
\end{equation}
Altogether, effective population size ($\Ne$), the number of loci ($\NbrLoci$) and the effect of a mutation ($a$), we can compute the variance of the environment ($\VarEnv$) that is required to reach a given heritability ($h^2$) as:
\begin{equation}
    \VarEnv = \VarGenetic \Multiply \left( \frac{1}{h^2} - 1 \right) = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply a^2 \Multiply \left( \frac{1}{h^2} - 1 \right) \label{eq:simu-var-env}
\end{equation}

>>>>

\title{Trait selection from within and between species variations while integrating genomics}
\rhead{\scshape Trait selection from within and between species variations}

@@

Even though the neutrality index can be applied for any trait, we so far assumed that several traits would to be independent of one another.
Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021}.
Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance-rejection), and the posterior distribution of $\CovarianceMatrix$ is sampled using a Gibbs sampler.
For each trait and each species, heritability ($\Heritability$) is set as a uniform distribution with user-defined boundaries.
The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.

@@

We analyzed a mammalian dataset, with the body mass and brain mass as quantitative traits.
The log-transformed values of body mass and brain mass are gathered from \textcite{tsuboi_breakdown_2018}.
We removed individuals not marked as adult, and split the data into males and females due to sexual dimorphism.
We discarded species with no intra-specific variability.
The mammalian nucleotide diversity is obtained from the Zoonomia project, with nucleotide divergence obtained on a set of neutral markers in \textcite{foley_genomic_2023}, and with nucleotide diversity measured as heterozygosity in \textcite{wilder_contribution_2023}.

@@

The inference framework was first tested on independently simulated datasets.
With the aim of applying the inference method to empirical datasets, the simulation parameters were chosen so as to match an empirically relevant mammalian empirical regime (see Materials and Methods).
First, we assumed a constant population size ($\Ne$) and a constant mutation rate ($\MutationRate$) across the phylogeny (fig.~\ref{fig:results-simulations}A\&B)
Simulations performed under stabilizing selection where detected as such ($\proba [\EstNI < 1] > 0.95$), no false negative was observed (blue in fig.~\ref{fig:results-simulations}A\&B).
Simulations performed under diversifying selection were detected as such ($\proba [\EstNI > 1] > 0.975$), no false negative was observed (red in fig.~\ref{fig:results-simulations}A\&B).
Simulations performed under neutral evolution were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$) in 77\% of the cases (yellow in fig.~\ref{fig:results-simulations}A\&B),while detected as stabilizing selection in 21\% of the cases, and detected as diversifying selection in 2\% of cases.
Second, for simulations under a fluctuating $\Ne$ and $\MutationRate$ (see fig.~\ref{fig:results-simulations}C\&D), simulations of diversifying selection was detected as such and stabilizing selection was detected as such in 100\% of the cases.
For simulations under neutral evolution, 51\% of the simulations were detected as such ($0.025 \leq \proba [\EstNI > 1] \leq 0.975$), 49\% of the simulations were detected as stabilizing selection ($\proba [\EstNI < 1] > 0.975$).

@@

\subsection*{Results on empirical data}\label{subsec:results-on-empirical-data}
For mammalian body mass, we obtained males (\Male) trait and genetic variation for $\NbrTaxa=36$ species, and in females (\Female) variation for $\NbrTaxa=26$ species.
The posterior estimate of neutrality index are $\EstNI_{\text{\Male}} = 0.34$ ($[0.22--0.52]$ for $95\%$ credible interval) and $\EstNI_{\text{\Female}} = 0.28$ ($[0.16--0.49]$ for $95\%$ CI).
In males and females, brain mass is under diversifying selection with associated posterior probabilities of $0.0$ and $0.0$ respectively.
Moreover, assuming that heritability ($\Heritability$) of body mass is uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection ($ \proba [\EstNI > 1]$) becomes $0.635$ in males and $0.914$ in females.
For mammalian brain mass, we obtained $\EstNI_{\text{\Male}} = 1.35$ ($[0.85--2.17]$ for $95\%$ CI, $\NbrTaxa=36$) and $\EstNI_{\text{\Female}} = 1.72$ ($[1.00--2.94]$ for $95\%$ CI, $\NbrTaxa=26$).
In males and females, brain mass is under diversifying selection with associated posterior probabilities of $0.877$ and $0.972$ respectively.
Moreover, assuming that heritability of brain mass is uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection becomes $1.0$ in both males and females.

For primates body mass, we obtained $\EstNI_{\text{\Male}} = 0.56$ ($[0.40,~0.78]$ for $95\%$ CI, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 0.39$ ($[0.28,~0.55]$ for $95\%$ CI, $\NbrTaxa=65$).
Assuming that heritability of body mass is uniformly distributed between 20\% and 40\%, posterior probabilities of diversifying selection becomes $1.0$ in males and $0.914$ in females.
For primates brain mass, we obtained $\EstNI_{\text{\Male}} = 1.95$ ($[1.40,~2.79]$ for $95\%$ CI, $\NbrTaxa=71$) and $\EstNI_{\text{\Female}} = 1.95$ ($[1.39,~2.79]$ for $95\%$ CI, $\NbrTaxa=65$).
In both males and females, brain mass is under diversifying selection, with associated posterior probabilities of $1.0$.
Importantly, these tests are invalid because the assumption of neutral genetic markers used for normalization is violated (see table~\ref{table:assumptions}).

@@

Heritability ($\Heritability$) is defined as:
\begin{equation}
    \Heritability = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}\label{eq:simu-heritability}
\end{equation}
Altogether, effective population size ($\Ne$), the number of loci ($\NbrLoci$) and the effect of a mutation ($a$), we can compute the variance of the environment ($\VarEnv$) that is required to reach a given heritability ($\Heritability$) as:
\begin{equation}
    \VarEnv = \VarGenetic \Multiply \left( \frac{1}{\Heritability} - 1 \right) = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply a^2 \Multiply \left( \frac{1}{\Heritability} - 1 \right) \label{eq:simu-var-env}
\end{equation}
